One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-10-25 doc/genesetdp.tex
Style changes.

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide}, some critique has been voiced concerning the completeness and rigour of current pathway databases. Hence, efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the terminology varies in the literature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.

% As the often used network randomization techniques are not completely random, such methods often introduce an undesirable bias in the statistics. For instance, the actual set of random networks is much larger than the ones one use in such test, namely the random networks that retain a certain topological characteristic\cite{newman2001random}.
% Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret, as such statistic in practice test a more complex null model than intended.
The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.
As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative. In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

We argue that the network perturbation methods used in most flavours of NEA is a red herring. Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.  The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.  In this question, the introduction of a network is not relevant. A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis. We argue that tests of significance in this type of analysis is naturally implemented as tests of the query set. That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of parametric score distributions for the calculations of significance.
Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.

>>>>

A limiting factor of ORA is the quality of the databases defining pathways. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide}, some critique has been voiced concerning the completeness and rigour of current pathway databases. Hence, efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. We will refer to such methods as {\em Network Enrichment Analysis} (NEA), although the concept also is known as crosstalk analysis.

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for network randomization, where the network definitions are randomly constructed by methods with varying degrees of topological conservation in respect to the original network.
The methods that are used for network randomization are often mentally untractable, as they include a big number of steps that are subject to complex constraints.

As an example, in the link assignment plus second-order conservation method \cite{mccormack2013statistical}, links are assigned at random, but sequentially and at each step they must follow constraints that depend on previous link assignments.
While this results in a network that is both randomized and contains some desired topological features, preserved from the original network, these nested constraints make it very hard to verify that the resulting network is representative.

In theory there is a space of all possible networks that could be made up from a certain number of nodes and node degree distribution, however, it is not clear that the link assignment and second-order conservation method samples evenly from that distribution.  And, if the resulting networks are not representative, the rejection of the resulting null hypothesis is uninformative, defeating the purpose of the statistical test.

We argue that the network perturbation methods used in most flavours of NEA are missleading. Users of NEA do not differ from users of ORA and want to test an association between a query set of genes and pathways.  The typical user has used an experimental or computational method to derive a set of query genes and wants to know if it is a random set, or if it is enriched with respect to a pathway.  In this question, the introduction of a network is not relevant. A network may help to increase the sensitivity of ORA methods, but for practical reasons it is better left out of the null hypothesis. We argue that tests of significance in this type of analysis is naturally implemented as tests of the query set. That is, is the query set a randomly selected set of genes or is there an over-representation of well-connected genes in the query set?

Hence, here we employ a different null model, more directly aimed at modelling the lack of association between query and pathway gene sets, which we formulate as ``there are not more links between the query and pathway gene sets than expected by chance''. We present a dynamic programming algorithm, which we dubbed GeneSetDP, that calculates the exact score distribution of any query of a given size. We also implemented one random sampling algorithm for the same task, that we called GeneSetMC. Both algorithms circumvent the need of network pertubations for the calculations of significance.
Combining the GeneSetDP and GeneSetMC with the scoring system of the popular method BinoX, we use simulations to demonstrate that both our algorithms produced unbiased statistics.
